Extended Early Adulthood and Attitude Change

### 2024-06-17 manuscript/formative_period_kk.Rmd
quality control, some edits to draft

Analyzing these questions requires data on attitude reports over time for members of different cohorts at different time points, presumably coming from the same population. While ideally we would be able to observe different cohorts responding to the same broad range of questions as they pass through the same age periods, we do not know of such a data source.

Instead, we focus on a range of questions tapping a similar domain -- national politics in the United States. We draw on data from seven different panels administered since the 1950s. These include four major panels administered by the American National Election Study in 1956-58-60, 1972-74-76, 1992-94-96, and 2000-02-04. These were all administered over a four-year period, with waves two years apart and contain a diverse array of attitudinal questions. We also use the General Social Survey's rotating panels from 2006-08-10, 2008-10-12, 2010-12-14. We also include the 1980 Major Panel study from the American National Election Study, a four-wave panel spanning about eight months. Finally, we incldue the the 1990-91-92 Merged File, a three-wave panel spanning two years. All seven surveys reflect nationally representative samples of U.S. households.

In all panels, we identified questions that tap several categories of political attitudes: (1) political efficacy and alienation; (2) attitudes toward social groups (such as labor unions, blacks, whites, the police, Democrats, liberals, etc.); (3) attitudes toward both domestic and policy issues; (4) ideological identification on a liberal-conservative scale; and (5) partisan identification. For the 1956-60, 1972-76, and 1980 panels, we broadly follow Krosnick and Alwin's selection of variables [-@krosnick1989]. In the 1992-96 panel, we follow Alwin's identification of questions [-@alwin2007]. For the 1990-92 panel, the 2000-04 panel, and the three GSS panels, we identify questions similar to those identified in the earlier panels.

This produces a set of 265 panel-attitude measures asked on a variety of scales tapping a diverse array of topics related to national government and politics in the United States. They include four- and five-point Likert scales, binary agree/disagree response scales, 100-point "feeling thermometers", and a range of other structures.

Most questions are asked in only one panel, but a small number of questions are replicated exactly across at least two panels (excluding questions duplicated across the GSS panels, where all questions are duplicated). There are theoretically relevant trade-offs from focusing only on the same questions over time. By focusing on the same questions over time, we control for question content and structure. If attributes of questions changes over time, that might affect the estimates of stability or reliability. However, the questions asked in major surveys change over time as the dominant issues change. By focusing only on questions that are repeated across eras, we limit our attention only to issues that have been salient for an extended period of time. These might behave different than the "issues of the day" -- they might get settled earlier in life, be more resistant to change, or be reported more reliably than other kinds of questions.

By focusing on all questions tapping the domains outlined above, the average level of stability and reliability reflect change in "important" issues -- including issues that remain important for a long time and issues that only maintain salience for a few years.

To ensure that our findings are robust to both sets of issues, we conduct a separate analysis of questions that are repeated across multiple panels (excluding questions replicated just across the GSS). These include partisan identification (asked in all panels), political ideology on a seven-point liberal-conservative scale (asked in \## panels), questions about federal spending on different priorities (asked in the 1990 and 2000 ANES panels and the GSS panels), and questions about government efficacy and alienation from the political process (asked in the 1970s, 1990s, and 2000 ANES panels), as well as others.

Weighting schemes are different across panels. When available, I use weights that account for differential selection. The 1972-76, 1980, and

## Methods of Analysis

There are a range of approaches to measure the amount of change a group undergoes. The starting point for our analysis if the Life Course Adaption Model (LCAM) formalized by Lersch [-@lersch2023]. The LCAM is a mixed-effects growth curve model that formalizes several key insights from the literature on attitude change over time. First, it assumes that survey respondents have an underlying disposition shaped by early-life socialization. Second, it assumes that people can, but do not necessarily, change this disposition over time as a result of aging and general life-course experiences. Third, it assumes that people fluctuate around their underlying dispositions over time as a result of random shocks and measurement error. Fourth, it assumes respondents can draw on their prior experiences when adapting to new events.

The model takes the form of:

$$ c_{i,t}^* = \beta_0 + \beta_ia_{i,t} + u_{0,i} + u_{i,t}a_{i,t} + \epsilon_{i,t}$$

where $c_{i,t}^*$ is the standardized survey response for respondent $i$ at time $t$, $\beta_0$ is the intercept with random term $u_{0,i}$ capturing early-life experiences. $\beta_1$ with random term $u_{i,t}a_{i,t}$ captures the individual-specific slope for age $a_{i,t}$. $\epsilon_{i,t}$ is random error around this individual-specific slope for person $i$ at time $t$.

By combining the main effect for time and the individual-level random effects for time, we generate a person-specific slope estimate, or the expected change in that attitude for a one-year change. Because we are unconcerned with whether respondents change in the same direction or tend to change in different directions (which could indicate either convergence or polarization, depending on where respondents start), the central quantity of interest for our analysis is the absolute change in respondents' latent trajectories for a year, or $|\beta_ia_{i,t} + u_{i,t}a_{i,t}|$. Averaging this quantity across all respondents in a group produces an average absolute change in responses for a year for members of that group.

As a second quantity of interest, we calculate the mean square residual for respondents in a particular group. Groups might differ both in their linear (systematic) change, as well as whether they are consistent at responding around that trajectory. High variance around a latent trajectory could indicate that respondents are either more error-prone in making responses or more subject to temporary considerations. To evaluate this, we calculate:

$$\frac{1}{n}\sum{(y_{i,t} - \hat{y_{i,t}})^2} $$

@@

@@

@@

>>>>

Our general question is whether, when presented with the same question during the same age range, more recent cohorts will demonstrate more change over time. Analyzing this question requires data on opinion and attitude reports over time for members of different cohorts at different time points, coming from the same general population. To construct such a data source, we combine several panel surveys conducted by the American National Election Studies and the General Social Survey spanning about 60 years. Both the ANES and GSS are designed to generalize to the non-institutionalized U.S. adult (18+) population, and can be thought of as tapping the same general population. We included a total of 12 panels in our search: ANES panels from 1956-60, 1972-76, 1980, 1990-92, 1992-97, 2000-04, 2016-20, and 2020-22; and GSS panels from 2006-10, 2008-12, 2010-14, and 2016-20. We evaluated an additional panel conducted by the ANES in 2008-09 (with an additional follow-up in 2010), but no questions overlapped this panel and any other panel.

For reasons explained below, we identified questions that were asked in at least three different panels with at least one of those panels asking some respondents to answer the question at three or more separate time points. Because three of the GSS panels overlap considerably (2006-10, 2008-12, and 2010-14), we did not treat these as separate panels for the purposes of generating three distinct time points. In other words, questions needed to appear in these panels and an ANES panel to be included in our analysis.

This produced a total of ### questions tapping a broad range of topics including political and ideological identification, government policy preferences, preferences for federal spending on different priorities, views on government efficacy, views on social issues and social change, and sentiment toward various political and social groups.

## Methods

When questions are observed in multiple panels, differences in the amount of change we observe could be attributable to either changes in the population's ability to report stable attitudes over time or to the structural features of the survey administration. Specifically, the amount of change observed over time might be due to variation in the duration between waves, with longer duration between observations leading to more wave-to-wave change. Across our panels, duration between questions ranges from a few days (for questions asked in both pre- and post-election surveys in the ANES) to about five years (questions asked in 1992 and 1997), with peaks around 2 years and 4 years.

In the one-item, one-panel case (e.g., partisan identification in the 1972-76 panel), we could model the absolute change from wave-to-wave as a function of the duration between observations.

$$ |y_{i,t}-y_{i,t-1} | = \beta_0 + \beta_1*d_i + \epsilon_i $$

In this equation, $d_{i}$ captures the duration of time between $t-1$ and $t$ in years. In the 1972-76 panel, respondents are asked their partisan identification three times: in the 1972 pre-election survey, in the 1974 pre-election survey, and the 1976 pre-election survey, each about two years apart. If a person participates in all three waves, they have three observations of $|y_{i,t}-y_{i,t-1} |$ and $d_i$: 1972 and 1974 (about 2 years), 1974 and 1976 (about 2 years), and 1972 and 1976 (about 4 years).

If respondents have a stable underlying partisan identification they report with error, and no real change occurs, then in expectation, duration should be unrelated to absolute change in responses over time, and $\beta_1 = 0$. In this formulation, $\beta_0$ captures the expected absolute change in responses over time when the duration between waves is 0, or, more intuitively, the random deviation between waves that is unrelated to durable change (random noise). Conversely, $\beta_1$ captures the expected expected change in absolute difference between waves for a one-unit (year) change in duration of observation, or the amount of absolute change we would expect over-and-above the change driven by random noise ($\beta_1$). For example, if a respondent shifts from "strong Democrat" (point 0 on the scale) to "Independent" (point 3 on the scale) between 1972 and 1974, and stays at "Independent" for the 1976 wave, they will have differences of 3, and 0 for the two-year duration (an average of 1.5) and 3 for the four-year duration, generating a positive coefficient for $\beta_1$ if this kind of switching behavior is common in the populations.

Our theoretical question deals primarily with whether these coefficients ($\beta_0$ and $\beta_1$) change over time. To explore that, we can add an additional term for time and interact the duration term with time.

$$ |y_{i,t}-y_{i,t-1} | = \beta_0 + \beta_1d + \beta_2t + \beta_3(d*t) + \epsilon_i $$

In this formulation, $\beta_2$ captures the change in the baseline level of random change over time. A negative value of $\beta_2$ would suggest that people are becoming more consistent reporters of their true underlying opinion over time. $\beta_3$ captures the expected change in absolute change over time for a one-year change in duration. A positive $\beta_3$ value would suggest that people are more likely to undergo a "real" change in their opinion over time in later panels than earlier panels.

A challenge for our analysis is that while the panels we explore often have relatively large samples, most questions are only asked in a few panels. As such, we lack large numbers of any particular question, meaning the time trend for any specific question is likely to be unstable. One potential solution is to pool questions and estimate a single time trend for stability in attitudes over time, but that is potentially problematic if the kinds of questions asked in later panels are systematically different from the kinds of questions asked in earlier panels. It could also obscure the fact that some items might be becoming more stable over time while others are becoming less stable.

Multilevel models provide an alternative solution to estimate variation in trends in the presence of uncertainty.

$$ |y_{i,j,t}-y_{i,j,t-1} | = \beta_{0j} + \beta_{1j}d + \beta_{2j}t + \beta_{3}(d*t) + \epsilon_i $$

Where:

$$ \beta_{0j} = \gamma_0 + \upsilon_{0j} $$
$$ \beta_{1j} = \gamma_1 + \upsilon_{1j} $$
$$ \beta_{2j} = \gamma_2 + \upsilon_{2j} $$
$$ \beta_{3j} = \gamma_3 + \upsilon_{3j} $$

In this framework, the first-level model models the absolute difference in a pair of responses to the same question given by the same respondent, similar to what was presented above. These equations are nested at the level of the item. Each item is assumed to have its own intercept (expected variance at time = 0 and duration = 0) and effects for time and duration. These effects are assumed to be normally distributed around a main effect. In practice, we see minimal variation in the question-specific coefficients for the interaction between duration and time, and as such do not allow these to vary at the question level.

For all models, we normalize response scales between 0 and 100.

@@

Model 1 in Table XXX presents the main effects and variance components of the equation presented above.

The intercept suggests that, on average, respondents varied by about a fifth of the scale (17 points), controlling for duration. This is a rather high level of variance, but consistent with previous studies that find that, on any particular question, survey respondents tend to make large wave-to-wave changes. Variance around this intercept is quite large, with a standard deviation of about 5.67, meaning many items vary by as much as 20 percent of the scale, while some vary by less than 10 points.

Second, duration is weakly positively associated with wave-to-wave absolute change. For every additional year between waves, respondents' absolute deviance increases about .71 points. Notably, variance around this main effect is relatively small (.43), meaning that two thirds of questions have duration effects that fall between .23 and 1.14. Broadly speaking, we can say that most wave-to-wave change is non-persisting, consistent with other recent work [@kiley2020; @lersch2023]. 
The main effect of year of observation in decades is also small (-.37), but negative. In other words, on average, respondents exhibit less wave-to-wave change in their responses the longer a question is asked. However, there is significantly more variance around this main effect, with a standard deviation of about .88. This suggests that while some items show decreased wave-to-wave change over time, others show increased wave-to-wave change over time.

@@

--
There are theoretically relevant trade-offs from focusing only on the same questions over time. By focusing only on the same questions over time, we control for question content and structure and ensure that the trend over time is due to something other than changes in survey administration. Because our panels span more than five decades that saw dramatic improvements in survey question design, questions included in more recent surveys might simply be of higher quality, with less wave-to-wave change, than items included in older surveys.

@@

--
